---
title: "College Enrollments in Illinois"
author: "Collin Jaeger"
date: "3/30/2018"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

```

## About the data
Source: IBHE Data Book

Table I-2: Total Fall Enrollments by Gender, Race or National Origin, and Type of Institution, and Level of Instruction

http://legacy.ibhe.org/IBHEDatabook/ChapterI/Table%20I-2.aspx

One file (.csv) was downloaded for each year (1996 - 2016, except 2013).

The data are very untidy . . . but at least they're *consistently* untidy!

## Load a few libraries
```{r}

library(tidyverse)
library(readxl)
library(vegan)

```

## Import and tidy the data
```{r}

setwd('~/Documents/R_Data/IBHE_Enrollments/Data')

year = 1996

file = paste0(year, '.csv')

df = read.csv(file)

df %>% tbl_df

# Remove the commas from the data
no.commas = function(x){
  dx = as.character(df[ , x])
  dx2 = as.numeric(gsub(',', '', dx))
  dx2
}

df2 = data.frame(df[, 1], sapply(2:23, no.commas))

names(df2) = names(df)

df2$Level.of.Instruction = as.character(df2$Level.of.Instruction)

df2 %>% tbl_df

categories = df2 %>%
  tbl_df %>%
  filter(grepl('^ [[:blank:]]*', Level.of.Instruction)) %>%
  select(Level.of.Instruction)

categories$rownum = grep('^ [[:blank:]]*', df2$Level.of.Instruction)

l1 = split(df2, cumsum(1:nrow(df2) %in% categories$rownum))

names(l1) = categories$Level.of.Instruction

df3 = bind_rows(l1, .id='Category')

df3 %>% tbl_df

rownum2 = grep('Undergraduate', df3$Level.of.Instruction) - 1

l2 = split(df3, cumsum(1:nrow(df3) %in% rownum2))

names(l2) = df3$Level.of.Instruction[c(1, rownum2)]

df4 = bind_rows(l2, .id='School')
df4$Year = year

df4 %>% tbl_df

df5 = df4 %>% 
  tbl_df %>%
  filter( ! is.na(Grand.Total)) %>%
  filter( ! grepl('Total', School, ignore.case = T)) %>%
  rename(Level = Level.of.Instruction) %>%
  filter(Level %in% c('Undergraduate', 'Graduate')) %>%
  mutate(Category = trimws(Category)) %>%
  select(Year, Category, School, everything(), 
         -X, -Total.M, -Total.F, -Grand.Total)

df5 %>% tbl_df

```

## So, that code works to tidy *one* year.
## Turn it into a *function*, and use it on a *different* year.
```{r, include=F}

clean.data = function(year){

  setwd('~/Documents/R_Data/IBHE_Enrollments/Data')

  file = paste0(year, '.csv')
  
  df = read.csv(file)
  
  # Remove the commas from the data
  no.commas = function(x){
    dx = as.character(df[ , x])
    dx2 = as.numeric(gsub(',', '', dx))
    dx2
  }
  
  df2 = data.frame(df[, 1], sapply(2:23, no.commas))
  
  names(df2) = names(df)
  
  df2$Level.of.Instruction = as.character(df2$Level.of.Instruction)
  
  categories = df2 %>%
    tbl_df %>%
    filter(grepl('^ [[:blank:]]*', Level.of.Instruction)) %>%
    select(Level.of.Instruction)
  
  categories$rownum = grep('^ [[:blank:]]*', df2$Level.of.Instruction)
  
  l1 = split(df2, cumsum(1:nrow(df2) %in% categories$rownum))
  
  names(l1) = categories$Level.of.Instruction
  
  df3 = bind_rows(l1, .id='Category')
  
  rownum2 = grep('Undergraduate', df3$Level.of.Instruction) - 1
  
  l2 = split(df3, cumsum(1:nrow(df3) %in% rownum2))
  
  names(l2) = df3$Level.of.Instruction[c(1, rownum2)]
  
  df4 = bind_rows(l2, .id='School')
  df4$Year = year
  
  df5 = df4 %>% 
    tbl_df %>%
    filter( ! is.na(Grand.Total)) %>%
    filter( ! grepl('Total', School, ignore.case = T)) %>%
    rename(Level = Level.of.Instruction) %>%
    filter(Level %in% c('Undergraduate', 'Graduate')) %>%
    mutate(Category = trimws(Category)) %>%
    select(Year, Category, School, everything(), 
           -X, -Total.M, -Total.F, -Grand.Total)
  
  df5
}

```

```{r}

clean.data(1997)

clean.data(2002)

```


## Now, tidy and consolidate all of the data.
```{r}

yrs = c(1996:2012, 2014:2016)

out.list = lapply(yrs, clean.data)

out.df = bind_rows(out.list)

out.df
  
```

## Make a few plots
```{r}

out.df %>%
  ggplot() +
  geom_bar(aes(Year, fill=Category), show.legend=F) +
  labs(y='Number of Schools') +
  facet_wrap(~Category, scales='free', ncol=2)

```



```{r}

out.df %>%
  mutate(Total = out.df %>% 
           select(contains(".")) %>% rowSums()) %>%
  filter(Level == 'Undergraduate') %>%
  ggplot() +
  geom_line(aes(Year, Total, group=School),
             show.legend=F, alpha=0.5, col='blue') +
  facet_wrap(~Category, scales='free', ncol=2)

```

```{r, fig.height=8, fig.width=8}

out.df %>%
  mutate(Total = out.df %>% 
           select(contains(".")) %>% rowSums()) %>%
  filter(Category == 'Public Universities:',
         Level == 'Undergraduate') %>%
  ggplot() +
  geom_line(aes(Year, Total, group=School), 
            show.legend=F, col='blue') +
  labs(y='Undergraduate Enrollment',
       title='Public Universities') +
  facet_wrap(~School, ncol=3) +
  theme(text = element_text(size=22))


```



```{r}

out.df %>%
  mutate(Total = out.df %>% 
           select(contains(".")) %>% rowSums()) %>%
  filter(School == 'Northern Illinois University',
         Level == 'Undergraduate') %>%
  ggplot() +
  geom_line(aes(Year, Total), col='blue') +
  labs(y='Undergraduate Enrollment')

```

```{r}

out.df.long = out.df %>%
  gather(Group, Enrollment, 
         -Year, -Category, -School, -Level) %>%
  mutate(Group = gsub('.M', '_M', Group),
         Group = gsub('.F', '_F', Group)) %>%
  separate(Group, into=c('Race', 'Sex'), 
           sep='_', extra='merge')

out.df.long

```

```{r}

out.df.long %>%
  filter(grepl('Northern Illinois', School),
         Level == 'Graduate') %>%
  ggplot() +
  geom_line(aes(Year, Enrollment, color = Sex)) +
  facet_wrap(~Race, ncol=3)

```


```{r}

out.df.long %>%
  filter(School == 'Northern Illinois University',
         Level == 'Undergraduate') %>%
  group_by(Year, Race) %>%
  summarize(n = sum(Enrollment)) %>%
  mutate(Proportion = prop.table(n)) %>%
  ggplot() +
  geom_area(aes(Year, Proportion, fill=Race, group=Race)) +
  scale_fill_brewer(palette = 'Spectral')

```
```{r}

out.df.long %>%
  filter(School == 'Northern Illinois University',
         Level == 'Undergraduate') %>%
  mutate(Race = fct_collapse(Race, 
                             Other = c('Alien',
                                       'Asian',
                                       'Hawaiian',
                                       'Indian',
                                       'Other',
                                       'Two.or.more.races'))) %>%
  group_by(Year, Race) %>%
  summarize(n = sum(Enrollment)) %>%
  ggplot() +
  geom_area(aes(Year, n/1000, fill=Race, group=Race),
            position='fill') +
  labs(y='Proportion of students') +
  scale_fill_brewer(palette = 'Spectral')

```

```{r}
out.df.long %>%
  filter(School == 'Northwestern University',
         Level == 'Undergraduate') %>%
  mutate(Race = fct_collapse(Race, 
                             Other = c('Alien',
                                       'Asian',
                                       'Hawaiian',
                                       'Indian',
                                       'Other',
                                       'Two.or.more.races'))) %>%
  group_by(Year, Race) %>%
  summarize(n = sum(Enrollment)) %>%
  ggplot() +
  geom_area(aes(Year, n/1000, fill=Race, group=Race)) +
  labs(y = 'Enrollment (thousands)') +
  scale_fill_brewer(palette = 'Spectral')

```

## Racial diversity within a school over time (M and F combined)
```{r, fig.height=6, fig.width=10}

library(vegan)

out.df2 = out.df %>%
  gather(Group, Enrollment, 
         -Year, -Category, -School, -Level) %>%
  mutate(Group = gsub('.M', '_M', Group),
         Group = gsub('.F', '_F', Group)) %>%
  separate(Group, into=c('Race', 'Sex'), 
           sep='_', extra='merge') %>%
  mutate(Race = fct_collapse(Race, 
                             Other = c('Alien',
                                       'Asian',
                                       'Hawaiian',
                                       'Indian',
                                       'Other',
                                       'Two.or.more.races'))) %>%
  group_by(Category, School, Year, Level, Race) %>%
  summarize(Enrollment = sum(Enrollment)) %>%
  filter(Level == 'Undergraduate') %>%
  spread(Race, Enrollment) %>%
  mutate(Total = Other + Black + Hisp. + White)


out.df2$Diversity = diversity(out.df2[ , 5:8], 'simpson')

out.df2 %>%
  filter(Category == 'Public Universities:') %>%
  ggplot() +
  geom_line(aes(Year, Diversity, group=School), col='blue') +
  facet_wrap(~School)

```

## Diversity and Enrollment (over time)
```{r}

out.df2 %>%
  filter(School == 'Northern Illinois University') %>%
  ggplot() +
  geom_point(aes(Year, Diversity, 
                 group=School, size=Total, col=Total))

```

```{r}

sch = "Northwestern University"

out.df2 %>%
  filter(School == sch) %>%
  ggplot() +
  geom_point(aes(Total, Diversity), show.legend=F) +
  geom_label(aes(Total, Diversity, label=Year, col=Year),
             show.legend=F) +
  scale_color_gradient(low = 'lightblue', high = 'darkblue') +
  labs(x='Undergraduate Enrollment', 
       y="Racial Diversity (Simpson's Index)",
       title=sch)

```


